---
title: "RJAMI - Fast computation of Conditional Mutual Information for ceRNA network analysis
"
author: "Markus List, Andrea Hornakova, Jilles Vreeken, Marcel H. Schulz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Purpose

## Loading the library

```{r}
library(RJAMI)
```


## Testing if rJava is available

RJAMI is a wrapper package that conveniently makes the functionality of JAMI available to R users.
This is realized with the help of the rJava package that allows Java objects to
be manipulated from R. To make sure you can use RJAMI you therefore need to 
make sure that Java (version >= 1.8) and the rJava R package are installed and 
properly configured to work together. To allow users to test if everything 
is set up correctly, we provide a test function:

```{r}
test_jvm()
```

If there is a problem you will receive an error message here and you need to
fix the installation. Typical problems are 

- Java is not installed. Browse the web for instructions specific for your operating system to install Java.
- rJava is not installed. Fix by ```install.packages("rJava")```
- Java is not properly configured. Under linux this can often be fixed by
```sudo update-alternatives --config java```. Another option is to set the environment variable JAVA_HOME correctly.
- rJava is not properly configured. This can often be fixed by
```sudo R CMD javareconf```

If Java and rJava still don't work correctly, we recommend looking for help on [StackOverflow](https://stackoverflow.com/search?q=rJava).

## Citation

Hornakova and List et al., JAMI - Fast computation of Conditional Mutual Information for ceRNA network analysis
 (manuscript in preparation)
