---
title: "RJAMI - Fast computation of Conditional Mutual Information for ceRNA network analysis
"
author: "Markus List, Andrea Hornakova, Jilles Vreeken, Marcel H. Schulz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Purpose 

JAMI is a java 8 software tool that enables its users to compute conditional 
mutual information for inferring competing endogenous RNA networks from matched
gene and miRNA expression data. A detailled description of JAMI can be found
on its [website](http://jami.readthedocs.io/en/latest/index.html). 
The purpose of RJAMI is to make the functionality of JAMI conveniently available
to R users to simplify usage and downstream analyses. 

## Quickstart

If  
  - you are already familiar with JAMI 
  - RJAMI is already set up correctly
  - you have prepared the necessary input files or R objects

There are only two things you need to know.
First, you can manipulate the JAMI command
line arguments via ```jami_settings()```, e.g.

```{r, eval = FALSE}
jami_settings(numberOfPermutations = 1e6)
```

Current settings are listed when no arguments are provided.

Second, you can call JAMI via

```{r, eval = FALSE}
jami(gene_miRNA_interactions = your_interactions,
                   gene_expr = your_gene_expr,
                   mir_expr = your_mir_expr,
                   output_file = your_desired_result_file)

```

This function accepts common R data objects or file paths where the target file
needs to be formatted in a tab delimited text format JAMI can understand. In 
the remainder of this documentation we will explain how to use RJAMI in detail.

## Installing RJAMI

RJAMI is currently only available on github. It can be conveniently installed 
through the devtools R package

```{r, eval=FALSE}
library(devtools)
install_github("SchulzLab/RJAMI")
```

If installation was successful the package can be loaded via
```{r}
library(RJAMI)
```


## Before we begin: testing if rJava is available

RJAMI is a wrapper package that conveniently makes the functionality of JAMI available to R users.
This is realized with the help of the rJava package that allows Java objects to
be manipulated from R. Before you can use RJAMI you therefore need to 
make sure that Java (version >= 1.8) and the rJava R package are installed and 
properly configured to work together. To allow users to test if everything 
is set up correctly, we provide a test function:

```{r}
test_jvm()
```

If there is a problem you will receive an error message here and you need to
fix the installation. Typical problems are 

- Java is not installed. Browse the web for instructions specific for your operating system to install Java.
- rJava is not installed. Fix by ```install.packages("rJava")```
- Java is not properly configured. Under linux this can often be fixed by
```sudo update-alternatives --config java```. Another option is to set the environment variable JAVA_HOME correctly.
- rJava is not properly configured. This can often be fixed by
```sudo R CMD javareconf```

If Java and rJava still don't work correctly, we recommend looking for help on [StackOverflow](https://stackoverflow.com/search?q=rJava).

## Manipulate Settings

JAMI comes with a set of command line arguments that allow users to set 
important parameters of the software with respect to input data, 
computation strategy, p-value handling etc. The meaning of the individual 
settings is explained in 
[JAMI's user manual](http://jami.readthedocs.io/en/latest/index.html#usage).
In RJAMI, we store these settings for convenience in a settings manager. Its
current configuration can be listed like this:

```{r}
jami_settings()
```

To change one of these settings you can provide new values via the same function, e.g.

```{r}
jami_settings(pvalueCutOff = 0.05)
```

## Input

First, let's load some toy data:
```{r}
load(system.file("extdata",
                 "test_data.RData",
                 package = "RJAMI"))
```


JAMI expects three types of input:
- gene expression
- matched miRNA expression
- gene-miRNA or gene-miRNA-gene interactions to be considered. 

For the latter, two input formats are accepted. In the triple format, each 
gene-miRNA-gene interaction is listed separately like this:

```{r}
head(gene_mir_interactions_triplets)
```


In the set format, we list for each gene possible miRNA interaction
partners. gene-miRNA-gene triplets are then inferred by JAMI on the fly based on
shared miRNAs. This can be represented as a list in R:

```{r}
head(gene_mir_interactions_set)
```

NOTE: If you want to use the set format you need to update the settings 
accordingly:

```{r, eval= FALSE}
jami_settings(tripleFormat = FALSE)
```

For gene and miRNA expression data you can choose between Bioconductor 
ExpressionSet, data frames or matrices. An example for gene expression:

```{r}
head(gene_expr[,1:3])
```

And an example for miRNA expression data:

```{r}
head(mir_expr[,1:3])
```

For all three input options you can also provide a file path. In this case, we
expect that the target files follow the file formats that JAMI understands. These
are detailled in the [user manual](http://jami.readthedocs.io/en/latest/index.html#input).

## Execution

Once all input data are prepared you can call the ```jami()``` function. With 
above toy data sets this would look like this:

```{r}
result <- jami(gene_miRNA_interactions = gene_mir_interactions_triplets,
                   gene_expr = gene_expr,
                   mir_expr = mir_expr)

```

Here, we use a temporary file for storing the results. Alternatively, you can
provide a file path of your choice to save the results for later use. These
results can then be imported using standard R code, e.g. ```read.table```.

The result of JAMI has two parts. The first part lists the results of all tested
gene-miRNA-gene triplets:

```{r}
head(result$result)
```

In the second part, results have been aggregated for all gene-gene pairs. To
obtain an aggregate p-value we use the Fisher method.

```{r}
head(result$aggregated)
```


## Plotting the network graph

The significant ceRNA interactions reported by JAMI give rise to a ceRNA interaction network that can be drawn using any of the graph visualization libraries in R. Here we illustrate how the igraph R package can be used:

```{r, message=FALSE, warning=FALSE,error=FALSE, fig.width = 10, fig.height = 10}
library(igraph)
edges <- graph_from_data_frame(result$aggregated, directed = TRUE)
plot(edges)
```

## Downstream analysis

In R and in particular in the Bioconductor ecosystem dozens of packages have
been developed for gene set analysis. 
We refer to [bioconductor](http://bioconductor.org/) for an overview.

## Final words

RJAMI has been developed to make JAMI more accessible. Should you encounter any
issues with it please consider creating an issue on our github repository. 
Finally, if you find JAMI and RJAMI useful consider citing our paper.

## Citation

Hornakova and List et al., JAMI - Fast computation of Conditional Mutual 
Information for ceRNA network analysis
 (manuscript under revision)
